Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-11 paper.Rnw
Updated results using Pruessner segmentations, partial re-write

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
jens_RIDs = read.csv('data/jens_RID.csv')

adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)

tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_RIDs$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
jens_xval_data  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data.csv.gz'))
jens_xval_mean  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data_mean.csv.gz'))

mb.best.kappa = mean(subset(jens_xval_mean, reg_method   == "ANTS" & 
                                  method.mb    == "Majority Vote" &
                                  templates.mb == 19 & 
                                  atlases      == 9)$k.mb)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% \begin{figure}
% <<ADN1-library-composition, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% # These composition datasheets describe the subjects and subject DX in the 
% # template and atlas libraries for each voting run
% atlas_comp = read.csv('data/a2a_ants/atlas_library_composition.csv')
% templ_comp = read.csv('data/a2a_ants/template_library_composition.csv')
% a2a.best = subset(all_data, atlases == 9 &
%                     templates.mb == 19 &
%                     reg_method=='ANTS' &
%                     method.mb =='Majority Vote')
% 
% cumulative_dist = 
%   ggplot(a2a.best, aes(x=k.mb)) + stat_ecdf() + scale_y_continuous(breaks=seq(0,1,0.1))
% 
% a2a.best = subset(a2a.best, k.mb > 0.75) # drop the guys that suck
% 
% a2a.best = merge(a2a.best, atlas_comp, by=c("timestamp","atlases"))
% a2a.best = merge(a2a.best, templ_comp, by.x=c("timestamp","templates.mb"), by.y=c("timestamp","templates"))
% a2a.best$templates_AD_percent = (a2a.best$templates_AD / 19) * 100
% a2a.best$templates_MCI_percent = (a2a.best$templates_MCI / 19) * 100
% a2a.best.melted = melt(a2a.best, measure.vars=c("templates_AD_percent", "templates_MCI_percent"))
% 
% k_by_composition = 
%   ggplot(a2a.best.melted, aes(x=value, y=k.mb, colour=DX)) + facet_grid(variable ~ DX) + 
%   geom_boxplot(aes(group=as.factor(value), colour=DX)) +  geom_smooth(method="lm") 
% 
% snt.unilateral.vols = read.csv('data/a2a_snt_volumes.csv')
% a2a.best = merge(a2a.best, snt.unilateral.vols)
% 
% # bland-altman plots
% bland_mb_overall = with(a2a.best, bland_altman_plot(volume.snt, volume.mb, DX) + 
%   geom_smooth(method="lm") + 
%   scale_y_continuous(breaks=seq(-500, 500, 50)))
% 
% bland_ma_overall = with(a2a.best, bland_altman_plot(volume.snt, volume.ma, DX) + 
%   geom_smooth(method="lm") + 
%   scale_y_continuous(breaks=seq(-500, 500, 50)))
% 
% # just the trials with a high percentage of MCI in the template library
% a2a.best.highMCI = subset(a2a.best, templates_MCI_percent > 50)
% 
% bland_mb_highMCI = with(a2a.best.highMCI, bland_altman_plot(volume.snt, volume.ma, DX) + 
%   geom_smooth(method="lm") + 
%   scale_y_continuous(breaks=seq(-500, 500, 50)))
% 
%   
% @
% \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(jens_xval_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_y_continuous(breaks=seq(0,1,0.05)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb{} segmentations relative to
  SNT \gs{} segmentations for 69 ADNI subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Error bars indicate
  standard error.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb{} vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb{} and multi-atlas segmentation is
  computed for each subject across all ten rounds of validation. Shown on the
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test
  comparing the distribution of variances at each parameter setting
  (atlas/template library size). Only points where \mb{} mean variability is lower
  than multi-atlas are shown. Dashed lines indicate a p-value of 0.05 and 0.01.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
jens_atlases = read.csv('data/jens_atlases.csv')
jens_xval_data  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data.csv.gz'))
jens_xval_mean  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data_mean.csv.gz'))
jens_xval_mean$reg_method = factor(jens_xval_mean$reg_method, c("ANTS", "ANIMAL"))
jens_xval_mean$method.mb = factor(jens_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))
mb.best.kappa = mean(subset(jens_xval_mean, reg_method   == "ANTS" & 
                                  method.mb    == "Majority Vote" &
                                  templates.mb == 19 & 
                                  atlases      == 9)$k.mb)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)

tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlasess$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(jens_xval_mean, (templates.mb*atlases) %% 2 == 1 & templates.mb < 21),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases), linetype=reg_method)) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(.~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=interaction(reg_method, as.factor(atlases)))) + 
  stat_summary(fun.y=mean,geom='point',
               aes(y=k.mb, weight=1, group=interaction(reg_method, as.factor(atlases)))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=interaction(reg_method, as.factor(atlases))), linetype='solid', width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_y_continuous(breaks=seq(0,1,0.01)) + 
  scale_colour_hue(name="Number of Atlases") +
  scale_linetype(name="Registration Method") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb{} segmentations relative to
  Pruessner \gs{} segmentations for 69 ADNI subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Error bars indicate
  standard error.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.0002,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb{} vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb{} and multi-atlas segmentation is
  computed for each subject across all ten rounds of validation. Shown on the
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test
  comparing the distribution of variances at each parameter setting
  (atlas/template library size). Only points where \mb{} mean variability is lower
  than multi-atlas are shown. Dashed lines indicate a p-value of 0.05 and 0.01.}
  \label{fig:ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
